---
title: "Compare Sites/Samples - PCA"
author: "Ava Hoffman"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)  # CRAN v2.0.0
library(caret)      # CRAN v7.0-1
library(factoextra) # CRAN v1.0.7
```

## Load Data

Select the data source:

```{r}
# Choose one: data_source <- c("local", "github", "google")
data_source <- "google"
```

```{r}
# Locate data according to param above
if (data_source == "local") {
  # Look in pwd for specific snapshots
  sites_full <- read_csv("data/snapshots/BioDIGS_sites_20250929.csv")
  soil_full <- read_csv("data/snapshots/BioDIGS_soil_20250929.csv")
} else if (data_source == "google") {
  # All from Google (need access)
  googlesheets4::gs4_auth(email = T)
  sites_full <- googlesheets4::read_sheet(
    "https://docs.google.com/spreadsheets/d/1Mbuym6GgG2B3VFCbjhYy4vaPCx1_tIOBlBOlT1iIHZE/edit?gid=476949426#gid=476949426",
    sheet = "SITE DATA"
  )
  soil_full <- googlesheets4::read_sheet(
    "https://docs.google.com/spreadsheets/d/1Mbuym6GgG2B3VFCbjhYy4vaPCx1_tIOBlBOlT1iIHZE/edit?gid=476949426#gid=476949426",
    sheet = "SOIL SAMPLES",
    col_types = "c"
  )
} else {
  # Public snapshot on github if not downloaded
  sites_full <- read_csv(
    "https://github.com/FredHutch/GDSCNsoilsites/raw/refs/heads/main/data/snapshots/BioDIGS_sites_20250929.csv"
  )
  soil_full <- read_csv(
    "https://github.com/FredHutch/GDSCNsoilsites/raw/refs/heads/main/data/snapshots/BioDIGS_soil_20250929.csv"
  )
}
```

## Data cleanup

Drop some extra metadata columns:

```{r}
# Select only necessary columns
sites <-
  sites_full %>%
  select(
    c(
      site_id,
      origin,
      gps,
      mgmt_type,
      pct_imp_2024,
      usda_plant_hardiness_2023,
      tree_canopy_cover_2023,
      CEC_landcover_2020,
      NLCD_landcover_2024,
      sandtotal_r,
      silttotal_r,
      claytotal_r,
      om_r,
      fao_gsoc,
      fao_oxygen_class,
      fao_salt_class,
      fao_nutrient_class,
      fao_clay_frac,
      fao_gravel_frac,
      fao_sand_frac,
      fao_ph,
      fao_water_capacity_class,
      elevation
    )
  ) %>%
  mutate(origin = case_when(
    # Lump Montgomery County samples as one place of origin
    origin %in% c("Derwood, MD", "Boyds, MD", "Germantown, MD", "Bethesda, MD") ~ "Montgomery Co., MD",
    TRUE ~ origin
  )) %>%
  mutate(
    # Group a management types (can't have just a single observation for PCA) - will be 1s and 2s
    mgmt_type = case_when(
      mgmt_type %in% c(
        "Botanical Garden",
        "Mowed to control kudzu, but other than that unmanaged area"
      ) ~ "Managed: mowed/tended lawns, yard, or park",
      mgmt_type == "not yet provided" ~ NA,
      TRUE ~ mgmt_type
    )
  )

# Select only necessary columns
soil <-
  soil_full %>%
  select(
    c(
      collection_date,
      site_id,
      sample_id,
      ends_with("EPA3051"),
      water_pH,
      OM_by_LOI_pct,
      ends_with("Mehlich3"),
      Est_CEC,
      Base_Sat_pct,
      P_Sat_ratio
    )
  )
```

Add `NA` and clean some values and classes:

```{r}
soil <-
  soil %>%
  mutate(across(everything(), ~ na_if(., "Not yet tested"))) %>%
  mutate(As_EPA3051 = case_when(As_EPA3051 %in% c("< 3.0", "<3.0") ~ "0", TRUE ~ As_EPA3051)) %>%
  mutate(Cd_EPA3051 = case_when(Cd_EPA3051 %in% c("< 0.2", "<0.2") ~ "0", TRUE ~ Cd_EPA3051)) %>%
  mutate(Cu_EPA3051 = case_when(Cu_EPA3051 %in% c("< 2.0", "<2.0") ~ "0", TRUE ~ Cu_EPA3051)) %>%
  drop_na() %>%
  mutate(across(4:27, as.numeric))

sites <-
  sites %>%
  tidyr::separate(gps, into = c("lat", "lon"), sep = ", ") %>% # Make gps coords two columns
  mutate(
    lat = as.numeric(lat),
    lon = as.numeric(lon),
    mgmt_type = as.numeric(as.factor(mgmt_type)), # Just ones and twos, no need to one hot encode
    usda_plant_hardiness_2023 = as.numeric(factor( # Convert hardiness zone to factor
      usda_plant_hardiness_2023,
      levels = c(
        "1a",
        "1b",
        "2a",
        "2b",
        "3a",
        "3b",
        "4a",
        "4b",
        "5a",
        "5b",
        "6a",
        "6b",
        "7a",
        "7b",
        "8a",
        "8b",
        "9a",
        "9b",
        "10a",
        "10b",
        "11a",
        "11b",
        "12a",
        "12b"
      )
    ))
  ) %>%
  drop_na()

# Prepare one-hot columns
CEC_landcover_2020_dummy <- dummyVars(" ~ .", data = sites[, "CEC_landcover_2020"])
CEC_landcover_2020_onehot <- data.frame(predict(CEC_landcover_2020_dummy, newdata = sites[, "CEC_landcover_2020"]))
NLCD_landcover_dummy <- dummyVars(" ~ .", data = sites[, "NLCD_landcover_2024"])
NLCD_landcover_onehot <- data.frame(predict(NLCD_landcover_dummy, newdata = sites[, "NLCD_landcover_2024"]))

# Remove and replace landcover type column with one-hot encoded
sites <-
  sites %>% 
  select(!c(CEC_landcover_2020, NLCD_landcover_2024)) %>% 
  bind_cols(CEC_landcover_2020_onehot) %>% 
  bind_cols(NLCD_landcover_onehot)
```

## Sample level/scale

What is the PCA pattern at the sample level?

```{r}
# Includes soil characteristics, gps coords, mgmt type, environment etc.
sample_level <- soil %>% left_join(sites, by = join_by(site_id))

# Drop non-numeric columns
sample_level_pca <-
  sample_level %>%
  mutate(label = stringr::str_remove(sample_id, "Soil_")) %>%
  column_to_rownames("label") %>%
  select(!c(origin, collection_date, site_id, sample_id))


# Run PCA
pca_out <- prcomp(sample_level_pca, scale = TRUE)
summary(pca_out)

# Scree Plot: % Variance explained by different PCs
eig_val <- get_eigenvalue(pca_out)
p1 <- fviz_eig(pca_out, col.var = "blue")

# Correlation of variables in each PC
pca_var <- get_pca_var(pca_out)
p2 <- ggcorrplot::ggcorrplot(pca_var$cos2[, 1:2]) + theme(legend.position = "none")

# PCA labeled by sample
p3 <- fviz_pca_ind(pca_out, repel = F)

# PCA colored by origin
p4 <- fviz_pca_ind(
  pca_out,
  label = "none",
  habillage = sample_level$origin,
  mean.point = F,
  palette = viridis::magma(n = length(unique(
    sample_level$origin
  )), end = 0.9)
) + ggtitle("PCA: Sample level soil and site characteristics") + theme(legend.title = element_blank())

# Save combined plot
cowplot::plot_grid(p4, p2, ncol = 1, rel_heights = c(8, 2))
ggsave(
  "figures/PCA_by_sample.png",
  height = 10,
  width = 8,
  dpi = "print"
)
```

## Site level, with soil

What is the PCA pattern at the site level?

```{r}
# Includes soil characteristics, gps coords, mgmt type, environment etc.
site_level <- soil %>% select(!c(collection_date, sample_id)) %>% group_by(site_id) %>% 
  summarize(across(everything(), mean)) %>% 
  left_join(sites, by = join_by(site_id))

# Drop non-numeric columns
site_level_pca <-
  site_level %>%
  column_to_rownames("site_id") %>%
  select(!c(origin))


# Run PCA
pca_out <- prcomp(site_level_pca, scale = TRUE)
summary(pca_out)

# Scree Plot: % Variance explained by different PCs
eig_val <- get_eigenvalue(pca_out)
p1 <- fviz_eig(pca_out, col.var = "blue")

# Correlation of variables in each PC
pca_var <- get_pca_var(pca_out)
p2 <- ggcorrplot::ggcorrplot(pca_var$cos2[, 1:2]) + theme(legend.position = "none")

# PCA labeled by site
p3 <- fviz_pca_ind(pca_out, repel = F)

# PCA colored by origin
p4 <- fviz_pca_ind(
  pca_out,
  label = "none",
  habillage = site_level$origin,
  mean.point = F,
  palette = viridis::magma(n = length(unique(
    site_level$origin
  )), end = 0.9)
) + ggtitle("PCA: Site level soil and site characteristics") + theme(legend.title = element_blank())

# Save combined plot
cowplot::plot_grid(p4, p2, ncol = 1, rel_heights = c(8, 2))
ggsave(
  "figures/PCA_by_site.png",
  height = 10,
  width = 8,
  dpi = "print"
)
```

## Site level, without soil, sampling dates

What is the PCA pattern at the site level, using mainly public data, with no soil characteristic data?

```{r}
# Sites only.. drop non-numeric columns
site_level_pca <-
  sites %>%
  rownames_to_column("drop") %>% 
  select(!drop) %>% 
  column_to_rownames("site_id") %>%
  select(!c(origin))

# Get clean var names
clean_names <- data.frame(longname = names(site_level_pca))
names_dic <- read_csv("resources/variable_names.csv")
all_clean_names <- clean_names %>% left_join(names_dic) %>% pull(shortname)
colnames(site_level_pca) <- all_clean_names

# Run PCA
pca_out <- prcomp(site_level_pca, scale = TRUE)
summary(pca_out)

# Scree Plot: % Variance explained by different PCs
eig_val <- get_eigenvalue(pca_out)
p1 <- fviz_eig(pca_out, col.var = "blue")

# Correlation of variables in each PC
pca_var <- get_pca_var(pca_out)
p2 <- ggcorrplot::ggcorrplot(pca_var$cos2[, 1:2]) + theme(legend.position = "none") + theme(axis.text.x = element_text(size = 8))

# PCA labeled by site
p3 <- fviz_pca_ind(pca_out, repel = F)

# PCA colored by origin
p4 <- 
  fviz_pca_ind(
    pca_out,
    label = "none",
    habillage = sites$origin,
    mean.point = F,
    palette = viridis::magma(n = length(unique(sites$origin)), end = 0.8)
  ) +
  ggtitle("PCA: Site level characteristics only (no soil measures)") +
  theme(legend.title = element_blank()) +
  scale_shape_manual(values = c(1:25, 1, 2))


# Save combined plot
cowplot::plot_grid(p4, p2, ncol = 1, rel_heights = c(8, 5))
ggsave(
  "figures/PCA_by_site_no_soil.png",
  height = 8,
  width = 10,
  dpi = "print"
)
```
